\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-04-14 paper/main.tex
working on text edits to address reviewer comments

Within the Python~\cite{vanR95} community, external packages that are
published in the most popular repositories~\cite{Pyth03, cond15} are
associated with version numbers and tags that allow users to guarantee
they are installing exactly the same code across different computing
environments~\cite{CoghStuf13}.  While it is \textit{possible} to
manually install the intended version of every dependency of a Python
script or package, manually tracking down those dependencies can
impose a substantial burden on the user and create room for mistakes
and inconsistencies. Further, when dependency versions are left
unspecified, replicating the original computing environment becomes
difficult or impossible.

@@

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the default notebook experience. Like standard
Jupyter notebooks, \texttt{davos}-enhanced notebooks allow researchers
to include text, executable code, and media within a single file. No
further setup or installation is required, beyond what is needed to
run standard Jupyter notebooks. And like virtual environments,
\texttt{davos} provides a convenient mechanism for fully specifying
(and installing, as needed) a complete set of Python dependencies,
including specific package versions.

@@

@@

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion
comment.'' An onion comment is a special type of inline comment that
may be placed on a line containing a \texttt{smuggle} statement to
customize how \texttt{davos} searches for the smuggled package locally
and, if necessary, downloads and installs it. Onion comments follow a
simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with
\texttt{davos} intuitive and familiar. To construct an onion comment,
users provide the name of the installer program (e.g., \texttt{pip})
and the same arguments one would use to manually install the package
as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when
installing it) may differ from its top-level module name (i.e., the
name used when importing it). In such cases, an onion comment may be
used to ensure that \texttt{davos} installs the proper package if it
cannot be found locally:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of the
installer's behavior, they provide a mechanism for precisely
controlling how, where, and when smuggled packages are installed.
Critically, if an onion comment includes a version
specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the
smuggled package exists locally, \texttt{davos} will extract its
version information from its metadata and compare it to the specifier
provided. If the two are incompatible (or no local installation is
found), \texttt{davos} will download, install, and load a suitable
version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to \texttt{smuggle} specific VCS references
(e.g., Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding
arguments to the installer program. In addition to preventing onion
comments from being used as a vehicle for shell injection attacks,
this enables \texttt{davos} to adapt its behavior based on how particular
flags will affect the behavior of the installer program. For example, if an
onion comment contains either the \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall} flag,
\texttt{davos} will not bother checking for a local copy of the smuggled
package before installing a new one:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, the \texttt{--no-input} flag will temporarily enable
\texttt{davos}'s non-interactive mode (see
Sec.~\ref{subsec:config}), and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

@@

@@

Since its initial release, \texttt{davos} has found use in a variety
of applications. In addition to managing computing environments for
multiple ongoing research studies, \texttt{davos} is being used by
both students and instructors in programming and methods courses such
as Storytelling with Data~\cite{Mann21d} (an open course on data
science, visualization, and communication) and Laboratory in
Psychological Science~\cite{Mann22} (an open course on experimental
and statistical methods for psychology research) to simplify
distributing lessons and submitting assignments, as well as in online
demos such as \texttt{abstract2paper}~\cite{Mann21e} (an example
application of GPT-Neo~\cite{GaoEtal20, BlacEtal21}) to share
ready-to-run code that installs dependencies automatically.

>>>>

Within the Python~\cite{vanR95} community, external packages that are
published in the most popular repositories~\cite{Pyth03, cond15} are
associated with version numbers and tags that allow users to guarantee
they are installing exactly the same code across different computing
environments~\cite{CoghStuf13}.  While it is \textit{possible} to
manually install the intended version of every dependency of a Python
script or package, manually tracking down those dependencies can
impose a substantial burden on the user and create room for mistakes
and inconsistencies. Further, when dependency versions are left
unspecified, replicating the original computing environment becomes
difficult or impossible~\citep{PimeEtal19}.

@@

We designed \texttt{davos} to occupy a ``sweet spot'' between these extremes.
\texttt{davos} is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
\texttt{davos}-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run standard Jupyter
notebooks. And like virtual environments, \texttt{davos} provides a convenient
mechanism for fully specifying (and installing, as needed) a complete set of
Python dependencies, including specific package versions, which are contained
and isolated from the rest of the user's system.

@@

Installing new packages in a notebook using standard approaches (e.g., system
commmands) affect the runtime environment. This could lead to undesired
behaviors. For example, running a notebook that installs new packages in the
user's primary system environment might alter their main system installation
and/or containerized environment in unexpected ways (e.g., changing package
versions, causing conflicts with other packages, etc.). To protect against
undesired changes to the runtime environment, \texttt{davos} incorporates its
own virtual environment for managing packages it installs. When \texttt{davos}
is imported, a new virtual environment (folder) is created automatically. The
folder's name may be customized to support multi-notebook projects. Any
\texttt{smuggle}d packages that were not available in the notebook's runtime
environment are installed to the current project folder. The runtime
environment remains unaffected by \texttt{davos}'s behavior (see
Sec.~\ref{subsec:project}).

@@

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion comment.'' An
onion comment is a special type of inline comment that may be placed on a line
containing a \texttt{smuggle} statement to customize how \texttt{davos}
searches for the smuggled package locally and, if necessary, downloads and
installs it. Onion comments follow a simple format based on the ``type
comment'' syntax introduced in PEP 484~\cite{vanREtal14}, and are designed to
make managing packages with \texttt{davos} intuitive and familiar. To construct
an onion comment, users provide the name of the installer program (e.g.,
\texttt{pip}) and the same arguments one would use to manually install the
package as desired via the command line: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1} \end{center} Occasionally,
a package's distribution name (i.e., the name used when installing it) may
differ from its top-level module name (i.e., the name used when importing it).
In such cases, an onion comment may be used to ensure that \texttt{davos}
installs the proper package if it cannot be found locally: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2} \end{center} Because onion
comments may be constructed to specify any aspect of the installer's behavior,
they provide a mechanism for precisely controlling how, where, and when
smuggled packages are installed. Critically, if an onion comment includes a
version specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific version
requested, or satisfies the given version constraints. If the smuggled package
exists locally, \texttt{davos} will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), \texttt{davos} will download, install, and
load a suitable version of the package instead: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3} \end{center} Onion
comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.): \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4} \end{center}
\texttt{davos} processes onion comments internally before forwarding arguments
to the installer program. In addition to preventing onion comments from being
used as a vehicle for shell injection attacks, this enables \texttt{davos} to
adapt its behavior based on how particular flags will affect the behavior of
the installer program. For example, if an onion comment contains either the
\texttt{-I}/\texttt{--ignore-installed}, \texttt{-U}/\texttt{--upgrade}, or
\texttt{--force-reinstall} flag, \texttt{davos} will not bother checking for a
local copy of the smuggled package before installing a new one: \begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5} \end{center} Similarly,
the \texttt{--no-input} flag will temporarily enable \texttt{davos}'s
non-interactive mode (see Sec.~\ref{subsec:config}). Installation flags that
affect the user's system outside of the current notebook are disabled by
default (see Sec.~\ref{subsec:projects}).

, and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

@@

\subsubsection{Projects}\label{subsec:projects}

Because \texttt{davos} can install new packages, running the code in a
\texttt{davos}-enhanced notebook might (in principle) affect the behavior of
\textit{other} Python-based software (e.g., other notebooks, scripts, etc.) by
altering which packages are installed in the runtime environment. This could
lead to undesired consequences. For example, suppose Person A develops a
notebook (Notebook A) for their research project. We will assume that Notebook
A does not use \texttt{davos} to manage project dependencies.  If Person A
runs a \texttt{davos}-enhanced Notebook B, e.g., sent by another developer,
might this unexpectedly affect the behavior of Notebook A?

We implemented a ``project'' system in \texttt{davos} to protect against the
above scenario. By default, importing \texttt{davos} creates a new project
folder in the user's home directory (contained within a hidden \texttt{.davos}
folder). The default project name is computed to uniquely identify each
notebook according to its filename and path. Any packages that were not
originally available in the notebook's runtime environment are installed to the
notebook's project directory. When external libraries are \texttt{smuggle}d,
\texttt{davos} temporally appends the current project directory to the search
path. Because the user's system path remains unchanged, and because none of the
runtime environment's packages are altered, the user's system and runtime
environment remain unaffected (aside from installing the \texttt{davos} package
itself to the runtime environment).

Each notebook's project may be customized by setting \texttt{davos.project} to
any string that can be used as a valid folder name in the user's operating
system. By customizing the project name, users can build multi-notebook
projects that share the same core set of dependencies without needing to
duplicate each package for each notebook in the project.

Finally, if the user \textit{does} wish to modify their runtime environment,
this may be done by setting \texttt{davos.project} to \texttt{None}. Doing so
will cause any packages installed by \texttt{davos} to affect the user's
runtime environment.  This is generally not recommended, as it can lead to
unintended consequences for other code that shares the runtime environment.

@@

Since its initial release, \texttt{davos} has found use in a variety of
applications. In addition to managing computing environments for multiple prior
and ongoing research studies~\citep{MannEtal23, OwenMann23, ZimaEtal23},
\texttt{davos} is being used by both students and instructors in programming
and methods courses such as Storytelling with Data~\cite{Mann21d} (an open
course on data science, visualization, and communication) and Laboratory in
Psychological Science~\cite{Mann22} (an open course on experimental and
statistical methods for psychology research) to simplify distributing lessons
and submitting assignments, as well as in online demos such as
\texttt{abstract2paper}~\cite{Mann21e} (an example application of
GPT-Neo~\cite{GaoEtal20, BlacEtal21}) to share ready-to-run code that installs
dependencies automatically.
